---
title: "Assignment 4 2019"
output: html_document
---

**John Stout**

ANCOVA and MIXED EFFECTS ANOVA ASSIGNMENT

Instructions: Respond to the questions below (use code chunks where necessary). For text responses, put your response in bold text. Make sure your document knits as HTML, but only upload the .rmd file.

There are two datasets involved in this assignment.

1.) Your first dataset is an ANCOVA example. The story is that you're testing three hangover cues. The "drink" variable codes for this. However, there is variation in how drunk your test subjects are. This is assessed in the "drunk" rating column. Your assessment is the "well" variable, which assesses how well the subject feels the morning after taking the hangover cure. 

Explain why ANCOVA might be an appropriate test in this case.

**In this situation, you have an independent variable "drink" that has three levels of hangover cures while also having a covariate "drunk" since individuals varying in their alcohol tolerance. Using an ANCOVA, you can control for the effects of the covariate on the dependent variable "well" responses to the hangover cure. Therefore, because we have an extra variable that could be influencing our dependent variable if ignored, controlling for its effect can help reveal true relationships. **

2.) Here is the data URL: http://www.cogneurolab.org/GradStats/A5_D01.csv
Load the data into R. (this was adapted from assignment 5 last year, that's not a typo)
```{r}
# loading data
data1_og = read.csv("http://www.cogneurolab.org/GradStats/A5_D01.csv")
```

3.) Name one assumption of ANCOVA, and then test that assumption in R.

**The relationship between the dependent variable and the covariate is linear.**

```{r}
library(tidyverse)

# separate dataframe
data1_sep = data1_og %>% separate(drink.well.drunk,c("drink","well","drunk"))

# make data numeric
data1_sep$drink = as.numeric(data1_sep$drink)
data1_sep$well  = as.numeric(data1_sep$well)
data1_sep$drunk = as.numeric(data1_sep$drunk)

# look at data
ggplot(data1_sep, aes(x=data1_sep$drunk, y=data1_sep$well)) + 
  geom_point() + geom_smooth(method=lm, se = FALSE, color = "darkred") + 
  theme_classic()

# Pearsons correlation
data1_cor = cor.test(x=data1_sep$drunk,y=data1_sep$well,method=c("pearson"))
p1_cor    = data1_cor[["p.value"]]
r1_cor    = data1_cor[["estimate"]]

```
**Above, I visualized and ran a correlation analysis to show that the relationship between the covariate "drunk" and dependent variable "well" is linear and negative (r = `r round(r1_cor,3)`, p = `r round(p1_cor,3)`.**

4.) Run an ANCOVA on this data in R using ezANOVA and write a description of the outcome.
```{r}
library(ez)
data1_sep$subject = as.factor(1:15)
data1_sep$drunk   = as.factor(data1_sep$drunk)
data1_sep$drink   = as.factor(data1_sep$drink)

data1_ancova = ezANOVA(
  data               = data1_sep,
  dv                 = well,
  wid                = subject,
  between            = drink,
  between_covariates = drunk,
  type               = 3
  )

```

5.) Your second dataset is a mixed-effects example. This is basically the beverage / imagery example we worked in class, but with gender as a between-subjects factor. To refresh your memory, participants watch beer, wine, and water commercials that contain negative, neutral, and positive imagery. They then rate the ads. Load the data. http://www.cogneurolab.org/GradStats/A5_D02.csv

```{r}
# load data
data2_og = read.csv("http://www.cogneurolab.org/GradStats/A5_D02.csv")
```

6.) Plot the data appropriately to the mixed-effects approach. There are multiple ways to do this, but it's usually best to plot the data in separate plots for the between-subjects factor in a mixed-effects ANOVA. Use standard error for error bars.

```{r}
# partial eta squared = SSn/(SSn+SSd)

# separate dataframe
list_sep = strsplit(c(colnames(data2_og)), ".", fixed = TRUE)

# add commas where needed
list_sep = list_sep[[1]]
list_sep[3]="beer.pos"; list_sep[4]="beer.neg"; list_sep[5]="beer.neut"; list_sep[6]="wine.pos";
list_sep[7]="wine.neg"; list_sep[8]="wine.neut"; list_sep[9]="water.pos"; list_sep[10]="water.neg";
list_sep[11]="water.neut"

# separate data
data2_sep = data2_og %>% separate(colnames(data2_og),list_sep)

# grabbed these from lecture 19 to help reformat
df_long = data2_sep %>% gather(condition, rating, beer.pos:water.neut) 
df_long = df_long %>% 
  separate(condition, c("Bev","Emo"))
df_long$Bev = as.factor(df_long$Bev)
df_long$Emo = as.factor(df_long$Emo)

# df for bar graph
df_long$rating = as.numeric(df_long$rating)
df_bar = df_long %>% 
  group_by(gender,Bev,Emo) %>%
  summarise_at(c("rating"),funs(mean,sem=sd(.)/sqrt(n())))

# it was not clear which gender is 1 and 2
ggplot(df_bar, aes(y = mean, x = Emo, fill=Bev))+
  geom_bar(stat="identity", color="black", position=position_dodge()) +
  geom_errorbar(aes(ymin = mean-sem, ymax = mean+sem), width=.2, position=position_dodge(.9)) +
  theme_classic() +
  ylab("Rating") + facet_grid(. ~ gender)

```

7.) Run a mixed-effects ANOVA in R. Hold off on describing / writing up this result. 
```{r}
library(ez)
df_long$Bev    = as.factor(df_long$Bev)
df_long$Emo    = as.factor(df_long$Emo)
df_long$gender = as.factor(df_long$gender)

# anova
df_anova   = ezANOVA(
  data     = df_long, 
  wid      = Participant, 
  dv       = rating, 
  within   = .(Bev, Emo),
  between  = gender,
  detailed = TRUE, 
  type     = 3
  )

#print(df_anova)
```

8.) Explain and then implement an appropriate approach this analysis following your mixed-effects ANOVA results.
**There was no 3 way interaction, no bev/emot interaction, no gender/emo interaction, but there was a gender/bev interaction. Furthermore, the gender/bev interaction had similar variances based on Mauchleys sphericity test, W = `r df_anova[["Mauchly's Test for Sphericity"]][["W"]][[2]]`, p = `r round(df_anova[["Mauchly's Test for Sphericity"]][["p"]][[2]],3)`, therefore, it is reasonable to do pairwise ttests between the levels of gender.**

```{r}
df_long2 = data2_sep %>% gather(condition, rating, beer.pos:water.neut) 

df_gen1   = df_long2 %>% filter(gender != 2)
df_gen2   = df_long2 %>% filter(gender != 1)

# pairwise t-test
df_gen1_pwise = pairwise.t.test(as.integer(df_gen1$rating), as.factor(df_gen1$condition), paired=TRUE, p.adjust.method="holm")

df_gen2_pwise = pairwise.t.test(as.integer(df_gen2$rating), as.factor(df_gen2$condition), paired=TRUE, p.adjust.method="holm")

# print data
#print(df_gen1_pwise$p.value)
#print(df_gen2_pwise$p.value)

```

9.) Write up your results for 7-8 in proper APA formatted using embedded R to print out statistics.
**To determine whether participants rated ads in a manner dependent on the imagery shown, we ran a mixed effects ANOVA with ratings as the within subject factors and gender as the between subject factor. We first found an interaction between gender and beverage (F(`r df_anova[["ANOVA"]][["DFn"]][[5]]`,`r df_anova[["ANOVA"]][["DFd"]][[5]]`) = `r df_anova[["ANOVA"]][["F"]][[5]]` p = `r round(df_anova[["ANOVA"]][["p"]][[5]],digits = 3)`, partial-eta squared = `r df_anova[["ANOVA"]][["SSn"]][[5]]/(df_anova[["ANOVA"]][["SSn"]][[5]]+df_anova[["ANOVA"]][["SSd"]][[5]])`). We therefore followed up with pairwise t-tests, separating by gender. With respect to gender 1, we found a significant difference between; beer.negative/water.neutral (p = `r round(df_gen1_pwise[["p.value"]][[4]],3)`), beer.negative/wine.negative (p = `r round(df_gen1_pwise[["p.value"]][[6]],3)`), beer.positive/water.neutral (p = `r round(df_gen1_pwise[["p.value"]][[20]],3)`), water.neutral/wine.positive (p = `r round(df_gen1_pwise[["p.value"]][[40]],3)`), wine.negative/wine.positive (p = `r round(df_gen1_pwise[["p.value"]][[56]],3)`), wine.neutral/wine.positive (p = `r round(df_gen1_pwise[["p.value"]][[64]],3)`). For gender 2, we found differences between beer.negative/water.positive (p = `r round(df_gen2_pwise[["p.value"]][[5]],3)`), beer.negative/wine.positive (p = `r round(df_gen2_pwise[["p.value"]][[8]],3)`), beer.neutral.water.positive (p = `r round(df_gen2_pwise[["p.value"]][[13]],3)`), beer.neutral/wine.negative (p = `r round(df_gen2_pwise[["p.value"]][[14]],3)`), beer.neutral/wine.neutral (p = `r round(df_gen2_pwise[["p.value"]][[15]],3)`), beer.neutral/wine.positive (p = `r round(df_gen2_pwise[["p.value"]][[16]],3)`), water.negative/wine.positive (p = `r round(df_gen2_pwise[["p.value"]][[32]],3)`), water.neutral/water.positive (p = `r round(df_gen2_pwise[["p.value"]][[37]],3)`), water.neutral/wine.negative (p = `r round(df_gen2_pwise[["p.value"]][[38]],3)`), water.neutral/wine.neutral (p = `r round(df_gen2_pwise[["p.value"]][[39]],3)`), water.neutral/wine.positive (p = `r round(df_gen2_pwise[["p.value"]][[40]],3)`), wine.negative/wine.positive (p = `r round(df_gen2_pwise[["p.value"]][[56]],3)`).**

**Taken together, individuals of gender 1 tend to prefer beer even when presented with negative or neutral ratings with respect to other beverages. However, when presented with positive imagery, they will prefer alcoholic beverages over water. With respect to gender 2, they tend to prefer wine when faced with any imagery, with beer and water tending to be similar in preference.**


